Matching Bullets

### 2015-12-08 imaging-paper.Rnw
Some small typos

options(replace.assign=TRUE,scipen=3, digits=2)
@

@@

%\hh{XXX add parameter estimates? $\hat{r} \approx 4500$ - indicating a 9 mm bullet, amount of degrees covered by land?  }
The scatterplot in Figure~\ref{fig:residual} shows the residuals of such a fit.
\hh{In this instance, the radius is estimated as $\hat{r} = \Sexpr{pars$radius}\mu m = \Sexpr{pars$radius/1000}mm$, which is consistent with a 9 mm bullet fired in a Ruger P-85. }
The residuals are dominated, as expected, by the grooves, which show up as large positive residuals. For a surface cross cut at $x = 100$ there is a residual circular structure that does not show up for all cross sections.

@@

\begin{figure}
\begin{subfigure}[t]{\linewidth}
\caption{\label{fig:loess_step1}Step 1 the Loess fit algorithm. For a fixed crosscut (x held constant at 243.75) height values for land 5 of bullet 1 are plotted across different values of y.}
<<loess_step1, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\linewidth', warning=FALSE,  message=FALSE>>=
br111_bullet <- get_bullet(paste(datadir,"Br1 Bullet 1-5.x3p", sep = "/"))
br121_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-1.x3p", sep = "/"))
br122_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-2.x3p", sep = "/"))
br123_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-3.x3p", sep = "/"))
br124_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-4.x3p", sep = "/"))
br125_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-5.x3p", sep = "/"))
br126_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-6.x3p", sep = "/"))

@@

Next, we fit a Loess regression to the data. Figure \ref{fig:loess_fit} provides a look at the Loess fit, in blue, overlaid against the processed image of land 5 of bullet 1. The fit seems to do a reasonable job of capturing the structure of the image. Figure \ref{fig:loess_resid} shows the residuals from this fit. It is clear that there is a rather significant amount of autocorrelation, which make sense given the presence of striations and markings. Certainly, tuning the parameters for this regression, such as the Loess span, could prove to do an even better job at making the markings more clear. \hh{XXX did you follow up on that? What is the effect of using a different span?}

\begin{figure}[hbtp]
\begin{subfigure}[t]{\linewidth}
\caption{\label{fig:loess_fit} Loess fit for land 5 of bullet 1.}
<<loess_fit, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\linewidth', warning=FALSE,  message=FALSE>>=

@@

my.loess$fitted
@
\end{subfigure}
\begin{subfigure}[t]{\linewidth}
\caption{\label{fig:loess_resid} Residuals of Loess fit for land 5 of bullet 1.}
<<loess_resid, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\linewidth', warning=FALSE>>=
my.loess$resid + ylab("Residual of Loess Fit")
@
\end{subfigure}
\caption{Fit and residuals of a loess fit to land 5 of bullet 1 (Barrel 1).}
\end{figure}

We can reproduce a similar figure to the one from the circular fits by overlaying the residuals from each loess fit for bullet 2 onto an image of the loess residuals from land 5 of bullet 1. This is shown in Figure~\ref{fig:loessfits}.

@@

\section{Application}
We have implemented a web interface to automate many of the steps outlined in the previous section. The web interface, developed using Shiny \cite{shiny}, allows a user to upload two x3p files representing scans of a bullet between two grooves. A 3D viewer, using Plotly \cite{plotly}, allows the user to rotate and zoom to examine surface features of the images. A screenshot of the application is shown in Figure~\ref{fig:app}. The application is available at the url \url{http://erichare.shinyapps.io/x3prproto}.

\begin{figure*}
\includegraphics[width=\textwidth]{images/app.png}
\caption{\label{fig:app}{Screenshot of the web interface for analyzing bullet images. \hh{The shot shows two lands from the Hamby study. Although the lands are a match, this is visually very hard to confirm.}}}
\end{figure*}

The mouse is the primary control input for interacting with the surface plots. One can mouse over the plots to view the coordinate values for each point. Clicking and dragging anywhere on the plot will allow rotation. Scrolling up and down allows the user to zoom in and out respectively. There are also a set of controls available at the top right of the page, which will allow the selection of some other rotation methods, taking of screenshots, and reseting the camera to the default view.

@@

The application includes a button which allows the automatic creation of a residual plot. An x coordinate value can be selected, and a cross-section of the two bullet at the x value will be taken. The previously described Loess regression procedure will be applied to the two cross sections. A residual plot will be produced beneath the surface plots allowing for an assessment of the similarity of the markings. It is intended that users will perform this procedure at different x values in order to see whether the markings indicate a match at different points along the bullet.

@@

\includegraphics[width=\columnwidth]{images/side-by-side-manual-match.png}
\caption{\label{fig:manualmatch}\hh{Manually adjusted side-by-side comparison of land 5 of bullet 1 and land 1 of bullet 2.}}
\end{figure}
\begin{figure}
\includegraphics[width=\columnwidth]{images/matchup-rgl.png}
\caption{\label{fig:manualmatch-rgl}\hh{Different view of the manually adjusted side-by-side comparison of land 5 of bullet 1 and land 1 of bullet 2.}}
\end{figure}

@@

ggplot() + 
  geom_rect(aes(xmin = xmin, xmax = xmax, fill=factor(type)), ymin = -6, ymax=6.5,  data = lines, alpha=0.2, show.legend = FALSE) +
  theme(legend.position="bottom") +
  geom_text(aes(x = meany), y= -5.5, label= "x", data = subset(lines, !match)) +
  geom_text(aes(x = meany), y= -5.5, label= "o", data = subset(lines, match)) +
  ylim(c(-6,6.5)) + theme_bw() +
  geom_line(data=lofX, aes(x=y, y=l30, group=bullet, colour=bullet)) +
  scale_colour_brewer("", palette="Set1") +
  scale_fill_brewer("", palette="Set2", na.value=alpha("grey60", 0.5)) +
    theme(legend.position = c(1,1.2), legend.justification=c(1,1),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab(expression(paste("Crosscut values (in ",mu,"m)", sep=""))) +
  xlab("y")
@
    }
\end{subfigure}
\caption{\label{fig:match}Matching striation marks: smooth (a), identify peaks and valley (b), and match peaks and valleys between crosscuts (c).}
\end{figure}

@@

\subsection{Automatic Matching Algorithm}
\hh{The automated matching routine is then as follows:
Identify a stable region for each bullet and pick the lowest crosscut level in this region (assumption here: lower crosscut levels show better expression of individual marks XXX need to still check)

@@

@@

\hh{XXX another idea for the rest of the paper: use some datamining methods to match the Hamby bullets }
Using recursive partitioning, we find a decision tree \cite{breiman:1984, rpart, rpart.plot} predicting matches based on features derived from the image files. The resulting tree is shown in Figure~\ref{fig:tree}. A total of 105 lands is being matched correctly. Interestingly, the number of consecutive matching striae is not featuring as an important variable in this evaluation. Instead, cross-correlation (ccf) between the crosscuts is very predominant. Aside from cross correlation, the number of matches (num.matches) and the cumulative height of the joint peaks (sumpeaks) between crosscuts is involved in the decision. Within matches and non-matches, the values of CMS and cross-correlation seem to be almost independent (see Figure~\ref{fig:ccfs-match}), but both are highly predictive in distinguishing matches from non-matches. Out of the two, cross-correlation has higher  predictive power. This  does not  contradict earlier findings emphasizing the value of CMS on visual assessments of bullet matches: in those papers, assessments were based on purely visual inspection of either actual bullets or 2d microscopic images of bullets.
Neither one of these methods allows for an assessment of cross-correlations. This is one of the benefits of switching to a digitalized version of the images that preserves the 3D surface structure.

\begin{figure}
<<tree, echo=FALSE, fig.width=7, fig.height=4, out.width='\\linewidth'>>=
library(rpart)
library(rpart.plot)
CCFs <- read.csv("data-35-25/bullet-stats.csv")
includes <- setdiff(names(CCFs), c("b1", "b2", "data", "resID", "id.x", "id.y"))
rp1 <- rpart(match~., CCFs[,includes])  # doesn't include cms at all !!!!

per <- rp1$frame$yval # predictive probability for each node in the tree
prp(rp1, extra = 101,  box.col=c("grey90", "orange")[as.numeric(per > 0.5)+1])

CCFs$pred <- predict(rp1)
@
\caption{\label{fig:tree}Decision tree of matching bullets based on recursive partitioning.}
\end{figure}

<<pred:table, echo=FALSE, results='asis'>>=
library(xtable)
xtable(xtabs(~I(pred>0.75)+match, data=CCFs))
@

@@

%' 
%' <<rforest, echo=FALSE>>=
%' library(randomForest)
%' rtrees <- randomForest(factor(match)~., data=CCFs[,includes], ntree=300)
%' qplot(1:nrow(rtrees$err.rate), OOB, geom="line", data=data.frame(rtrees$err.rate)) + theme_bw() + ylim(c(0,NA))
%' CCFs$forest <- predict(rtrees, type="prob")[,2]
%' importance(rtrees)
%' qplot(pred, forest, geom="jitter", data=CCFs, colour=match)
%' @
%'

@@

\hh{Things we discussed}
\begin{enumerate}
\item the Hamby study only includes 35 bullets - this is not a particularly realistic scenario for automatic matching. Adding more non-matches into the mix is going to make things more realistic. 
\item The matching algorithm is sensitive to the parameter choices made. The distance between crosscuts (currently 25$\mu m$) to evaluate stability, as well as the cross-correlation factor (currently 0.9) we set as a minimum are affecting the final outcome. Another parameter is the amount of smoothing in identifying peaks and grooves (currently a window of 57.8375$\mu m$ is used, corresponding to a window of 17 values to the left and the right of an observation). While we tried to lay out in the paper the impact each of the parameter choices has on the matching, we are still far from an optimized scenario.

\item involve multiple crosscuts for each bullet in a match?
\item assessing different crosscuts at the same bullet allows us to assess a land's quality in terms of how stable it is -- generally, break-off patterns are less stable. }
\item we get probabilities for matches. 
\end{enumerate}

>>>>

options(replace.assign=TRUE,scipen=3)
@

@@

\hh{XXX add parameter estimates? $\hat{r} \approx 4500$ - indicating a 9 mm bullet, amount of degrees covered by land?  }
The scatterplot in Figure~\ref{fig:residual} shows the residuals of such a fit. The residuals are dominated, as expected, by the grooves, which show up as large positive residuals. For a surface cross cut at $x = 100$ there is a residual circular structure that does not show up for all cross sections.

@@

\begin{figure}
\begin{subfigure}[t]{\linewidth}
\caption{\label{fig:loess_step1}Step 1 the Loess fit algorithm. For a fixed crosscut (x held constant at 243.75) height values for part 1 of bullet 1 are plotted across different values of y.}
<<loess_step1, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\linewidth', warning=FALSE,  message=FALSE>>=
br111_bullet <- get_bullet(paste(datadir,"Br1 Bullet 1-5.x3p", sep = "/"))
br121_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-1.x3p", sep = "/"))
br122_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-2.x3p", sep = "/"))
br123_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-3.x3p", sep = "/"))
br124_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-4.x3p", sep = "/"))
br125_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-5.x3p", sep = "/"))
br126_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-6.x3p", sep = "/"))

@@

Next, we fit a Loess regression to the data. Figure \ref{fig:loess_fit} provides a look at the Loess fit, in blue, overlaid against the processed image of part 1 of bullet 1. The fit seems to do a reasonable job of capturing the structure of the image. Figure \ref{fig:loess_resid} shows the residuals from this fit. It is clear that there is a rather significant amount of autocorrelation, which make sense given the presence of striations and markings. Certainly, tuning the parameters for this regression, such as the Loess span, could prove to do an even better job at making the markings more clear. \hh{XXX did you follow up on that? What is the effect of using a different span?}

\begin{figure}[hbtp]
\begin{subfigure}[t]{\linewidth}
\caption{\label{fig:loess_fit} Loess fit for land 1 of bullet 1.}
<<loess_fit, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\linewidth', warning=FALSE,  message=FALSE>>=

@@

my.loess$fitted
@
\end{subfigure}
\begin{subfigure}[t]{\linewidth}
\caption{\label{fig:loess_resid} Residuals of Loess fit for land 1 of bullet 1.}
<<loess_resid, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\linewidth', warning=FALSE>>=
my.loess$resid + ylab("Residual of Loess Fit")
@
\end{subfigure}
\caption{Fit and residuals of a loess fit to land 5 of bullet 1 (Barrel 1).}
\end{figure}

We can reproduce a similar figure to the one from the circular fits by overlaying the residuals from each loess fit for bullet 2 onto an image of the loess residuals from part 1 of bullet 1. This is shown in Figure~\ref{fig:loessfits}.

@@

\section{Application}
We have implemented a web interface to automate many of the steps outlined in the previous section. The web interface, developed using Shiny \cite{shiny}, allows a user to upload two x3p files representing scans of a bullet between two grooves. A 3D viewer, using Plotly \cite{plotly}, allows the user to rotate and zoom to examine surface features of the images. A screenshot of the application is available in Figure \ref{fig:app}. The application is available at the url \url{http://erichare.shinyapps.io/x3prproto}.

\begin{figure*}
\includegraphics[width=\textwidth]{images/app.png}
\caption{\label{fig:app}{Screenshot of the web interface for analyzing bullet images.}}
\end{figure*}

The mouse is the primary control input for interacting with the surface plots. One can mouse over the plots to view the coordinate values for each point. Clicking and dragging anywhere on the plot will allow rotation. Scrolling up and down allows the user to zoom out and and respectively. There are also a set of controls available at the top right of the page, which will allow the selection of some other rotation methods, taking of screenshots, and reseting the camera to the default view.

@@

The application includes a button which allows the automatic creation of a residual plot from the Loess fits of two bullets. An x coordinate value can be selected, and a cross-section of the two bullet at the x value will be taken. The previously described Loess regression procedure will be applied to the two cross sections. A residual plot will be produced beneath the surface plots allowing for an assessment of the similarity of the markings. It is intended that users will perform this procedure at different x values in order to see whether the markings indicate a match at different points along the bullet.

@@

\includegraphics[width=\columnwidth]{images/side-by-side-manual-match.png}
\caption{\label{fig:manualmatch}\hh{Manually adjusted side-by-side comparison of land five of bullet 1 and land 1 of bullet 2.}}
\end{figure}
\begin{figure}
\includegraphics[width=\columnwidth]{images/matchup-rgl.png}
\caption{\label{fig:manualmatch-rgl}\hh{Different view of the manually adjusted side-by-side comparison of land five of bullet 1 and land 1 of bullet 2.}}
\end{figure}

@@

ggplot() + 
  geom_rect(aes(xmin = xmin, xmax = xmax, fill=factor(type)), ymin = -6, ymax=6.5,  data = lines, alpha=0.2, show.legend = FALSE) +
  theme(legend.position="bottom") +
  geom_text(aes(x = meany), y= -5.5, label= "x", data = subset(lines, !match)) +
  geom_text(aes(x = meany), y= -5.5, label= "o", data = subset(lines, match)) +
  ylim(c(-6,6.5)) + theme_bw() +
  geom_line(data=lofX, aes(x=y, y=l30, group=bullet, colour=bullet)) +
  scale_colour_brewer("", palette="Set1") +
  scale_fill_brewer("", palette="Set2", na.value=alpha("grey60", 0.5)) +
    theme(legend.position = c(1,1), legend.justification=c(1,1),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab(expression(paste("Crosscut values (in ",mu,"m)", sep=""))) +
  xlab("y")
@
    }
\end{subfigure}
\caption{\label{fig:match}Matching striation marks: smooth (a), identify peaks and valley (b), and match peaks and valleys between crosscuts (c).}
\end{figure}

@@

\hh{The automated matching routine is then as follows:
Identify a stable region for each bullet and pick the lowest crosscut level in this region (assumption here: lower crosscut levels show better expression of individual marks XXX need to still check)

@@

\hh{XXX needs more work: how does crosscut impact the results?  For the same bullet: how far apart can crosscuts be to still result in a match?}

@@

\hh{XXX another idea for the rest of the paper: use some datamining methods to match the Hamby bullets }
Using recursive partitioning, we find a decision tree \cite{breiman:1984, rpart, rpart.plot} predicting matches based on features derived from the image files. The resulting tree is shown in Figure~\ref{fig:tree}. A total of 105 lands is being matched correctly. Interestingly, the number of consecutive matching striae is not featured in this evaluation. Instead, cross-correlation (ccf) between the crosscuts is featured heavily. Aside from cross correlation, the number of matches between crosscuts is involved in the decision. Within matches and non-matches, the values of CMS and cross-correlation seem to be almost independent (see Figure~\ref{fig:ccfs-match}), but both are highly predictive in distinguishing matches from non-matches. Out of the two, cross-correlation has higher  predictive power. This  does not  contradict earlier findings emphasizing the value of CMS on visual assessments of bullet matches: in those papers, assessments were based on purely visual inspection of either actual bullets or 2d microscopic images of bullets.
Neither one of these methods allows for an assessment of cross-correlations. This is one of the benefits of switching to a digitalized version of the images that preserves the 3D surface structure.

\begin{figure}
<<tree, echo=FALSE, fig.width=7, fig.height=4, out.width='\\linewidth'>>=
library(rpart)
library(rpart.plot)
CCFs <- read.csv("data-35/bullet-stats.csv")
includes <- setdiff(names(CCFs), c("b1", "b2", "data", "resID", "id.x", "id.y"))
rp1 <- rpart(match~., CCFs[,includes])  # doesn't include cms at all !!!!

per <- rp1$frame$yval # predictive probability for each node in the tree
prp(rp1, extra = 101, fallen.leaves=TRUE, box.col=c("grey90", "orange")[as.numeric(per > 0.5)+1])

CCFs$pred35 <- predict(rp1)
@
\caption{\label{fig:tree}Decision tree of matching bullets based on recursive partitioning.}
\end{figure}

<<pred:table, echo=FALSE, results='asis'>>=
library(xtable)
xtable(xtabs(~I(pred35>0.5)+match, data=CCFs))
@

@@

%' 
%' <<rforest, echo=FALSE>>=
%' library(randomForest)
%' rtrees <- randomForest(factor(match)~., data=CCFs[,includes], ntree=1000)
%' CCFs$forest <- predict(rtrees)
%' importance(rtrees)
%' @
%'

@@

\hh{Things we discussed}
\begin{enumerate}
\item the Hamby study only includes 35 bullets - this is not a particularly realistic scenario for automatic matching. Adding more non-matches into the mix is going to make things more realistic. 
\item parameter optimization is going to drive some of the matches
\item involve multiple crosscuts for each bullet in a match?
\end{enumerate}
